---
title: "Data Review"
author: "Bruce R"
date: "2/1/2022"
output: html_document
---

```{r setup, include=FALSE}

####Make sure you load any required packages.

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Import Your Data

In the following code hunk, import your data.

```{r data_import, message=FALSE}

## build code dictionary for: tax codes <-> readable name
theColNames <- readxl::read_xls("data/STC_Historical_2020/STC_Historical_DB (2020).xls",
                                #skip = 1,
                                #na = "X",
                                n_max = 1,
                                col_names = FALSE)
the_codes <- readxl::read_xls("data/STC_Historical_2020/STC_Historical_DB (2020).xls",
                              skip = 1,
                              #na = "X",
                              n_max = 1,
                              col_names = FALSE)

codes_dict <- as.list(the_codes)
names(codes_dict) <- theColNames[,5:36]

rm(the_codes)
rm(theColNames)

## read data
df_original <- readxl::read_xls("data/STC_Historical_2020/STC_Historical_DB (2020).xls",
                       #skip = 2,
                       na = "X")

#  Wrangle
df <- df_original[2:3815,] %>% 
  filter(Year>=2005) %>% 
  select(-c(State, `FY Ending Date`)) %>% 
  mutate(State = str_sub(Name, 1,2)) %>% 
  select(-Name) %>% 
  filter(State != "US")
  # select categories, not the subcategories

# fix class of columns
df[,2:33] <- lapply(df[,2:33], parse_number)
df$State <- as.factor(df$State)

# convert to millions of dollars, original is thousands of dollars
df[,2:33] <- df[,2:33]/1000


# Make table for State Populations in 2010 and 2020
## data from https://www.census.gov/data/tables/time-series/dec/popchange-data-text.html
# direct link https://www2.census.gov/programs-surveys/decennial/2020/data/apportionment/apportionment.csv
pop_data <- read_csv("data/apportionment.csv")
pop_data <- pop_data %>% 
  filter(Year == 2010 | Year == 2020, `Geography Type` == "State") %>% 
  select(Name, Year, `Resident Population`) %>% 
  rename(Population = `Resident Population`,
         StateName = Name) %>% 
  filter(StateName %in% state.name)
pop_data$State <- as.factor(state.abb[match(pop_data$StateName, state.name)])

```

## Part 1

<!-- Using words, describe the visualization you are going to make using which variables/characteristics in your data: -->

<!-- *Example*: For my first figure, I am going to create a scatterplot that plots vehicle weight on the x axis and miles per gallon the y axis. I will create a two column tibble with these data. -->

<!-- In the code chunk below, show your work filtering the data and create the subset of data you will display graphically. -->
In the first figure, I will display the break up of a particular states income from the major tax categories. I will use a geom_area to show how the break down changes over time.  The code below shows the selection for the state Massachusetts. In the final project the state will be selected by the user.

First I take only the columns that are sums of subcategories of taxes (begin with "C"). I then remove "C105"(Total tax) which is the sum of all the categories. I then make it into a "long" data frame where the tax category is viewed as a variable. Finally I filter for the state Massachusetts.

```{r data_1}
df_major_cat <- df %>% 
  select(Year, State, starts_with("C")) %>% 
  #remove the total column
  select(-C105)
# also make a long data frame
df_major_cat_long <- df_major_cat %>% 
  pivot_longer(C107:C130, names_to = "TaxCode", values_to = "Millions_dollars", values_drop_na = TRUE) %>% 
  #add readable description of tax code
  mutate(TaxName = as.character(codes_dict[TaxCode]))

fig1_dat <- filter(df_major_cat_long, State == "MA")
####make sure you call the data so it will display in your report
fig1_dat
```

## Part 2

In the second figure, I will draw a chorpleth of a certain category of tax for the whole United States. In the code below, I use the year 2020 and tax T16 (tobacco tax), but in the final project, these will both be selectable.

First I take only the columns that are subcategories of taxes (begin with "T"). I then make it into a "long" data frame where the tax category is viewed as a variable. Finally I filter for the state Massachusetts.

```{r data_2}
# subcategories
df_sub_cat <- df %>% 
  select(Year, State, starts_with("T"))
# also make a long data frame
df_sub_cat_long <- df_sub_cat %>% 
  pivot_longer(T01:T99, names_to = "TaxCode", values_to = "Millions_dollars", values_drop_na = TRUE) %>% 
  #add readable description of tax code
  mutate(TaxName = as.character(codes_dict[TaxCode]))

fig_dat2<-df_sub_cat_long %>% filter(Year == 2020, TaxCode == "T16")
####make sure you call the data so it will display in your report
fig_dat2
```


## Part 3

<!-- Using words, describe the *third* visualization you are going to make using which variables/characteristics in your data: -->

<!-- *Example*: For the third figure, I will display a density plot of the quarter mile time for six cylinder cars. -->
In the third figure, I will make a dumbbell chart for the changes in per capita tax income change from 2010 to 2020 for a state. In the code below, I use Massachusetts, but in the final project, the state will be selectable.

First I take only the columns that are subcategories of taxes (begin with "T"). I then make it into a "long" data frame where the tax category is viewed as a variable. I then make a left join with the population data. This will only uuse the years 2010 and 2020 (the years of censuses). I then calculate a per capita column and convert(from millions of dollars) the money to true dollars and round to whole number. Finally I filter for the state Massachusetts make the year into a factor.



```{r data_3}
# subcategories
df_sub_cat <- df %>% 
  select(Year, State, starts_with("T"))
# also make a long data frame
df_sub_cat_long <- df_sub_cat %>% 
  pivot_longer(T01:T99, names_to = "TaxCode", values_to = "Millions_dollars", values_drop_na = TRUE) %>% 
  #add readable description of tax code
  mutate(TaxName = as.character(codes_dict[TaxCode]))



fig3_dat <- inner_join(pop_data, df_sub_cat_long)
fig3_dat <- fig3_dat %>% 
  mutate(perCapitaTax = round(Millions_dollars/Population*1000000)) %>% 
  filter(State == "MA")
fig3_dat$Year <- as.factor(fig3_dat$Year)
fig3_dat$TaxName <- as.factor(fig3_dat$TaxName)

####make sure you call the data so it will display in your report
fig3_dat
```

